Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-05-29 talk/dls2014/paper/paper.tex
add jython to the mix
more text in the performance section

\remi{For performance we first look at no-JIT behaviour of STM. Since
we cannot compete well even with CPython, we later show JIT benchmarks
where we see the unstable performance but also that we can still scale.
(with more work we can use our STM system to parallelise jitted code
too)} See figure \ref{fig:performance-nojit}

% TODO: pypy-nostm, Jython?
\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
  \caption{Comparing runtime between interpreters without a JIT\label{fig:performance-nojit}}
\end{figure}

% TODO: Jython, compare to cpython? or just jython as common baseline with no-jit?
\remi{Some benchmarks (figure \ref{fig:performance-jit} with enabled
JIT show that we can be competitive with the other solutions. It also
shows that more work is needed in that area to make performance more
stable.}

@@

>>>>

In this section we want to look at the general performance of our
system. As explained above, the JIT does not simply speed up the
performance by a certain factor. In order to better understand the
behaviour of our system, we therefore split this section into two
parts. We first look at how it behaves without the JIT, and then with
the JIT. We look at six benchmarks:

\begin{itemize}
\item \emph{btree} and \emph{skiplist}, which are both inserting,
  removing, and finding elements in a data structure
\item \emph{threadworms}, which simulates worms walking on a grid in
  parallel and checking for collisions with each other
\item \emph{mandelbrot}, \emph{raytrace}, and \emph{richards}, which
  all perform simple, independent computations in parallel
\end{itemize}

We use coarse-grained locking for the first three benchmarks and no
locking for the last three. This is important because Jython, which
uses fine-grained locking instead of a GIL, is only expected to scale
with the number of threads for the latter group. It is not able to
scale when using coarse-grained locking. STM, however, uses atomic
blocks instead, so it may still be able to scale since they are
implemented as simple transactions.

% To isolate factors we look at performance w/o JIT and perf w JIT.
% w/o JIT:
%  - it scales
%  - slower than reference CPython
%  - gil scales negatively
% w JIT:
%  - it sometimes still scales (more work needed)
%  - changed working set size because generally ~50x faster
%  - competing w our own JIT on non-stm a challenge
%  - gil scales negatively

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in \ref{fig:performance-nojit}.

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling. We also see Jython scale when we
expect it to (mandelbrot, raytrace, richards), and behave similar to
the GIL interpreters in the other cases.

PyPy using our STM system (pypy-stm-nojit) scales in all benchmarks to
a certain degree. We see that the average overhead from switching from
GIL to STM is \remi{$35.5\%$}, the maximum in richards is
\remi{$63\%$}. pypy-stm-nojit beats pypy-nojit already on two threads;
however, it never even beats CPython, the reference implementation of
Python. This means that without the JIT, our performance is not
competitive. We now look at how well our system works when we enable
the JIT.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
  \caption{Comparing runtime between interpreters without a JIT\label{fig:performance-nojit}}
\end{figure}

\paragraph{JIT benchmarks:} The speedups from enabling the JIT in
these benchmarks range from $10-50\times$. This is why we had to do
without Jython and CPython here, since they would be much further up
in the plots. Also, in order to get more stable results, we increased
the input size of all benchmarks to get reasonable execution times.

The results are shown in \ref{fig:performance-nojit}. We see that the
performance is much less stable. There is certainly more work required
in this area. In general, we see that the group of non-locked
benchmarks certainly scales best. The other three scale barely or not
at all with the number of threads. The slowdown factor from GIL to STM
ranges around \remi{$1-2.4\times$} and we beat GIL performance in half
of the benchmarks.

@@

Overall PyPy needs the JIT in order for its performance to be
competitive.  It would be interesting to see how using our STM system
in CPython would perform, but it is a lot of work. On its own, our
system scales well so we hope to also see that with the JIT in the
future.
